---
title: "African Epidemic Service - Biostatistics"
title-block-banner: true
format: 
    html:
        theme: flatly
        toc: true
        toc-depth: 4
        
execute:
    warning: false
    echo: true
    output: true
    
editor: visual
---

# DAY 1

## Set Up

### [cantstopwontstop.github.io/biostat_aes](https://cantstopwontstop.github.io/biostat_aes/)

[R - Cran](https://cran.r-project.org/){target="'_blank'"}

[RStudio](https://posit.co/downloads/){target="'_blank'"}

## Introduction

### RStudio

-   Files

    -   Script

    -   Quarto

-   Console

-   Environment

-   Plots

-   Packages

### R

#### Operators

##### Arithmetic Operators

Addition operator (+)

```{r}
200 + 175
```

Subtraction operator (-)

```{r}
200 - 175
```

Multiplication operator (\*)

```{r}
amount <- 200 * 175
amount
```

Division operator (/)

```{r}
200 / 175
```

Power operator (\^)

```{r}

200 ^ 5
```

Modulo operator (%%)

```{r}
10 %% 2
```

```{r}
10 %% 4
```

##### Relational Operators

Less than (\<)

```{r}
 
10 < 8
```

Greater than (\>)

```{r}
10 > 8
```

Equal to (==)

```{r}
10 == 8
```

Less than or equal to (\<=)

```{r}
10 <= 8
```

Greater than or equal to (\>=)

```{r}

10 >= 8
```

```{r}
8 >= 8
```

Not equal to (!=)

```{r}
10 != 8
```

##### Logical Operators

AND operator (&)

```{r}
TRUE & TRUE
TRUE & FALSE
FALSE & FALSE
```

OR operator (\|)

```{r}
TRUE | TRUE
TRUE | FALSE
FALSE | FALSE
```

NOT operator (!)

```{r}
!FALSE
!TRUE
```

#### Data Types

Logical

```{r}
status <- TRUE
status
```

Integer / Numeric

```{r}
age <- 48
age
```

Character

```{r}
name <- "Traore"
name
```

Vectors

```{r}
superstars <- c('Messi', 'Ronaldo', 'Salah', 'Mbappe', 'Saka', 'Dembele')
superstars
```

```{r}
ages <- c(38, 39, 33, 26, 23, 28)
ages
```

```{r}
world_cup_winner <- c(TRUE, FALSE, FALSE, TRUE, FALSE, TRUE)
world_cup_winner
```

Data frames

```{r}
superstar_data <- data.frame(superstars, ages, world_cup_winner)
superstar_data
```

#### Functions

```{r}
avg_age <- mean(superstar_data$ages)
avg_age
```

```{r}
oldest <- max(superstar_data$ages)
oldest
```

```{r}
youngest <- min(superstar_data$ages)
youngest
```

# DAY 2

## Load Packages

```{r}
library(tidyverse)
library(readxl)
library(DT)
```

## Practice Data

[Low Birth Weight](https://github.com/CantStopWontStop/biostat_aes/blob/master/Low%20birth%20weight.xlsx)

[Sierra Leone Ebola](https://github.com/CantStopWontStop/biostat_aes/blob/master/Ebola_Sierra%20Leone.xlsx)

## Load Data

```{r}
ebola_sl <- read_excel('Ebola_Sierra Leone.xlsx')
ebola_sl

```

## Key Functions

### summary()

```{r}
summary(ebola_sl)
```

### Pipe Function %\>%

```{r}
ebola_sl %>% summary() 
```

### head()

```{r}
ebola_sl %>% head(n = 20)
```

### tail()

```{r}
ebola_sl %>% tail()
```

### select()

Select is the function used for selecting columns

```{r}
ebola_demographics <- ebola_sl %>% select(age, sex, status)
ebola_demographics
```

### filter()

Filter is used to select the rows to keep or filter out rows

```{r}
ebola_sl %>% filter(age < 18 )
```

```{r}
ebola_sl %>% 
    filter(age < 18 & district == "Bo" )
```

The pipe function is used to chain operations. The output of the previous operation is fed in as the first argument in the following function using the pipe %\>%.

```{r}
ebola_sl %>% 
    filter(age < 18 & district == "Bo" ) %>% 
    select(age, sex, status, district)
```

### arrange()

```{r}
ebola_sl %>% 
    filter(age < 18 & district == "Bo" ) %>% 
    select(age, sex, status, district) %>% 
    arrange(age)
```

```{r}
ebola_sl %>% 
    filter(age < 18 & district == "Bo" ) %>% 
    select(age, sex, status, district) %>% 
    arrange(desc(age))
```

```{r}
ebola_sl %>% 
    filter(age < 18 & district == "Bo" ) %>% 
    select(age, sex, status, district) %>% 
    arrange(age %>% desc())
```

### mutate()

Mutate is used to create new columns or change existing columns.

```{r}
ebola_sl %>% 
    mutate(day_of_sample = as.Date(date_of_sample),
          day_of_onset = as.Date(date_of_onset), 
          timeframe = day_of_sample - day_of_onset) %>%
    select(day_of_sample, day_of_onset, timeframe)
    
```

### factor()

```{r}
ebola_sl_cleaned <- ebola_sl %>% 
    select(sex, status, date_of_onset, district, age) %>% 
    mutate(
        sex = factor(sex, 
                     levels = c(1,2),
                     labels = c("female", "male")),
        status = factor(status,
                        levels = c(1,0),
                        labels = c("positive", "negative") ),       # Convert to factor
        # Create 'age_group' based on the 'age' column
        age_group = case_when(
            age < 5 ~ "0-4", 
            age >= 5 & age < 15 ~ "5-14",
            age >= 15 & age < 25 ~ "15-24",
            age >= 25 & age < 45 ~ "25-44",
            age >= 45 & age < 65 ~ "45-64",
            age >= 65 ~ "65+",
            TRUE ~ NA_character_  # Handle missing values by assigning NA
        ),
        # Convert 'age_group' to factor with specified order
        age_group = factor(age_group, 
                           levels = c("0-4", 
                                      "5-14", 
                                      "15-24", 
                                      "25-44", 
                                      "45-64", 
                                      "65+")
                           )

    )

ebola_sl_cleaned %>% 
    datatable()
```

### summarise()

```{r}
ebola_sl %>% 
    summarise(
    mean_age = mean(age),
    max_age  = max(age),
    num_cases = sum(status)
    )
```

### drop_na()

Drop_na() is used to remove missing data. It can be done for the entire data frame or for specific columns.

##### Specific Columns

```{r}
ebola_sl %>% 
    drop_na(age) %>% 
    summarise(
    mean_age = mean(age),
    max_age  = max(age),
    num_cases = sum(status)
    )
 
```

##### All Columns

```{r}
ebola_sl %>% 
    drop_na() %>% 
    summarise(
    mean_age = mean(age),
    max_age  = max(age),
    num_cases = sum(status)
    )
```

### group_by()

Group by is used to make calculations/aggregations or changes by a group.

```{r}
ebola_sl_cleaned %>% 
    drop_na() %>% 
    group_by(sex) %>% 
    summarise(
        mean_age = mean(age),
        max_age  = max(age),
        min_sample_date = min(date_of_onset),
        max_sample_date = max(date_of_onset)
    )
```

```{r}
ebola_sl_cleaned %>% 
    drop_na() %>% 
    group_by(district) %>% 
    summarise(
        mean_age = mean(age),
        min_age  = min(age),
        max_age  = max(age),
        min_sample_date = min(date_of_onset),
        max_sample_date = max(date_of_onset)
    ) %>% 
    datatable()
```

## Plotting

ggplot()

aes()

geom\_\*()
