---
title: "African Epidemic Service - Biostatistics"
title-block-banner: true
format: 
    html:
        theme: flatly
        toc: true
        toc-depth: 4
        
execute:
    warning: false
    echo: true
    output: true
    
editor: visual
---

# DAY 1

## Set Up

### [cantstopwontstop.github.io/biostat_aes](https://cantstopwontstop.github.io/biostat_aes/)

[R - Cran](https://cran.r-project.org/){target="'_blank'"}

[RStudio](https://posit.co/downloads/){target="'_blank'"}

## Introduction

### RStudio

-   Files

    -   Script

    -   Quarto

-   Console

-   Environment

-   Plots

-   Packages

### R

#### Operators

##### Arithmetic Operators

Addition operator (+)

```{r}
200 + 175
```

Subtraction operator (-)

```{r}
200 - 175
```

Multiplication operator (\*)

```{r}
amount <- 200 * 175
amount
```

Division operator (/)

```{r}
200 / 175
```

Power operator (\^)

```{r}

200 ^ 5
```

Modulo operator (%%)

```{r}
10 %% 2
```

```{r}
10 %% 4
```

##### Relational Operators

Less than (\<)

```{r}
 
10 < 8
```

Greater than (\>)

```{r}
10 > 8
```

Equal to (==)

```{r}
10 == 8
```

Less than or equal to (\<=)

```{r}
10 <= 8
```

Greater than or equal to (\>=)

```{r}

10 >= 8
```

```{r}
8 >= 8
```

Not equal to (!=)

```{r}
10 != 8
```

##### Logical Operators

AND operator (&)

```{r}
TRUE & TRUE
TRUE & FALSE
FALSE & FALSE
```

OR operator (\|)

```{r}
TRUE | TRUE
TRUE | FALSE
FALSE | FALSE
```

NOT operator (!)

```{r}
!FALSE
!TRUE
```

#### Data Types

Logical

```{r}
status <- TRUE
status
```

Integer / Numeric

```{r}
age <- 48
age
```

Character

```{r}
name <- "Traore"
name
```

Vectors

```{r}
superstars <- c('Messi', 'Ronaldo', 'Salah', 'Mbappe', 'Saka', 'Dembele')
superstars
```

```{r}
ages <- c(38, 39, 33, 26, 23, 28)
ages
```

```{r}
world_cup_winner <- c(TRUE, FALSE, FALSE, TRUE, FALSE, TRUE)
world_cup_winner
```

Data frames

```{r}
superstar_data <- data.frame(superstars, ages, world_cup_winner)
superstar_data
```

#### Functions

```{r}
avg_age <- mean(superstar_data$ages)
avg_age
```

```{r}
oldest <- max(superstar_data$ages)
oldest
```

```{r}
youngest <- min(superstar_data$ages)
youngest
```

# DAY 2

## Packages

```{r}
library(tidyverse)
library(readxl)
```

## Practice Data

[Low Birth Weight](https://github.com/CantStopWontStop/biostat_aes/blob/master/Low%20birth%20weight.xlsx)

[Sierra Leone Ebola](https://github.com/CantStopWontStop/biostat_aes/blob/master/Ebola_Sierra%20Leone.xlsx)
