---
title: "African Epidemic Service - Biostatistics"
title-block-banner: true
format: 
    html:
        theme: flatly
        toc: true
        toc-depth: 4
        
execute:
    warning: false
    echo: true
    output: true
    
editor: visual
---

# DAY 1

## Set Up

### [cantstopwontstop.github.io/biostat_aes](https://cantstopwontstop.github.io/biostat_aes/)

[R - Cran](https://cran.r-project.org/){target="'_blank'"}

[RStudio](https://posit.co/downloads/){target="'_blank'"}

```{r}
library(tidyverse)
library(readxl)
library(DT)
```

## Introduction

### RStudio

-   Files

    -   Script

    -   Quarto

-   Console

-   Environment

-   Plots

-   Packages

### R

#### Operators

##### Arithmetic Operators

Addition operator (+)

```{r}
200 + 175
```

Subtraction operator (-)

```{r}
200 - 175
```

Multiplication operator (\*)

```{r}
amount <- 200 * 175
amount
```

Division operator (/)

```{r}
200 / 175
```

Power operator (\^)

```{r}

200 ^ 5
```

Modulo operator (%%)

```{r}
10 %% 2
```

```{r}
10 %% 4
```

##### Relational Operators

Less than (\<)

```{r}
 
10 < 8
```

Greater than (\>)

```{r}
10 > 8
```

Equal to (==)

```{r}
10 == 8
```

Less than or equal to (\<=)

```{r}
10 <= 8
```

Greater than or equal to (\>=)

```{r}

10 >= 8
```

```{r}
8 >= 8
```

Not equal to (!=)

```{r}
10 != 8
```

##### Logical Operators

AND operator (&)

```{r}
TRUE & TRUE
TRUE & FALSE
FALSE & FALSE
```

OR operator (\|)

```{r}
TRUE | TRUE
TRUE | FALSE
FALSE | FALSE
```

NOT operator (!)

```{r}
!FALSE
!TRUE
```

#### Data Types

Logical

```{r}
status <- TRUE
status
```

Integer / Numeric

```{r}
age <- 48
age
```

Character

```{r}
name <- "Traore"
name
```

Vectors

```{r}
superstars <- c('Messi', 'Ronaldo', 'Salah', 'Mbappe', 'Saka', 'Dembele')
superstars
```

```{r}
ages <- c(38, 39, 33, 26, 23, 28)
ages
```

```{r}
world_cup_winner <- c(TRUE, FALSE, FALSE, TRUE, FALSE, TRUE)
world_cup_winner
```

Data frames

```{r}
superstar_data <- data.frame(superstars, ages, world_cup_winner)
superstar_data
```

```{r}
superstar_data
```

```{r}
superstar_data %>% 
    mutate(year_of_birth = 2025 - ages,
           world_cup_win_yn = if_else(world_cup_winner == TRUE, "Yes", "No"))
```

```{r}
superstar_data %>% 
    mutate(year_of_birth = 2025 - ages,
           world_cup_win_yn = if_else(world_cup_winner, "Yes", "No"))
```

```{r}
superstar_data %>% 
    mutate(superstars = if_else(superstars == "Salah", "Egyptian King", superstars))
```

#### Functions

```{r}
avg_age <- mean(superstar_data$ages)
avg_age
```

```{r}
oldest <- max(superstar_data$ages)
oldest
```

```{r}
youngest <- min(superstar_data$ages)
youngest
```

# DAY 2

## Practice Data

[Low Birth Weight](https://github.com/CantStopWontStop/biostat_aes/blob/master/Low%20birth%20weight.xlsx)

[Sierra Leone Ebola](https://github.com/CantStopWontStop/biostat_aes/blob/master/Ebola_Sierra%20Leone.xlsx)

## Load Data

```{r}
ebola_sl <- read_excel('Ebola_Sierra Leone.xlsx')
ebola_sl

```

## Key Functions

### summary()

```{r}
summary(ebola_sl)
```

### Pipe Function %\>%

```{r}
ebola_sl %>% summary() 
```

### head()

```{r}
ebola_sl %>% head(n = 20)
```

### tail()

```{r}
ebola_sl %>% tail()
```

### select()

Select is the function used for selecting columns

```{r}
ebola_demographics <- ebola_sl %>% select(age, sex, status)
ebola_demographics
```

### filter()

Filter is used to select the rows to keep or filter out rows

```{r}
ebola_sl %>% filter(age < 18 )
```

```{r}
ebola_sl %>% 
    filter(age < 18 & district == "Bo" )
```

The pipe function is used to chain operations. The output of the previous operation is fed in as the first argument in the following function using the pipe %\>%.

```{r}
ebola_sl %>% 
    filter(age < 18 & district == "Bo" ) %>% 
    select(age, sex, status, district)
```

### arrange()

```{r}
ebola_sl %>% 
    filter(age < 18 & district == "Bo" ) %>% 
    select(age, sex, status, district) %>% 
    arrange(age)
```

```{r}
ebola_sl %>% 
    filter(age < 18 & district == "Bo" ) %>% 
    select(age, sex, status, district) %>% 
    arrange(desc(age))
```

```{r}
ebola_sl %>% 
    filter(age < 18 & district == "Bo" ) %>% 
    select(age, sex, status, district) %>% 
    arrange(age %>% desc())
```

### mutate()

Mutate is used to create new columns or change existing columns.

```{r}
ebola_sl %>% 
    mutate(day_of_sample = as.Date(date_of_sample),
          day_of_onset = as.Date(date_of_onset), 
          timeframe = day_of_sample - day_of_onset) %>%
    select(day_of_sample, day_of_onset, timeframe, date_of_onset, date_of_sample)
    
```

```{r}
ebola_sl %>% 
    mutate( timeframe = date_of_sample - date_of_onset) %>% 
    rename(time_frame = timeframe)
```

```{r}
ebola_sl %>% 
    mutate(date_of_sample = as.Date(date_of_sample),
          date_of_onset = as.Date(date_of_onset), 
          timeframe = date_of_sample - date_of_onset)
```

### factor()

```{r}
ebola_sl_cleaned <- ebola_sl %>% 
    select(sex, status, date_of_onset,date_of_sample, district, age) %>% 
    mutate(
        gender = factor(sex, 
                     levels = c(1,2),
                     labels = c("female", "male")),
        ebola_status = factor(status,
                        levels = c(1,0),
                        labels = c("positive", "negative") ),       # Convert to factor
        # Create 'age_group' based on the 'age' column
        age_group = case_when(
            age < 5 ~ "0-4", 
            age >= 5 & age < 15 ~ "5-14",
            age >= 15 & age < 25 ~ "15-24",
            age >= 25 & age < 45 ~ "25-44",
            age >= 45 & age < 65 ~ "45-64",
            age >= 65 ~ "65+",
            TRUE ~ NA_character_  # Handle missing values by assigning NA
        ),
        # Convert 'age_group' to factor with specified order
        age_group = factor(age_group, 
                           levels = c("0-4", 
                                      "5-14", 
                                      "15-24", 
                                      "25-44", 
                                      "45-64", 
                                      "65+")
                           )

    )

ebola_sl_cleaned %>% 
    datatable()
```

### summarise()

```{r}
ebola_sl %>% 
    summarise(
    mean_age = mean(age),
    max_age  = max(age),
    num_cases = sum(status)
    )
```

### drop_na()

Drop_na() is used to remove missing data. It can be done for the entire data frame or for specific columns.

##### Specific Columns

```{r}
ebola_sl %>% 
    drop_na(age) %>% 
    summarise(
    mean_age = mean(age),
    max_age  = max(age),
    num_cases = sum(status)
    )
 
```

##### All Columns

```{r}
ebola_sl %>% 
    drop_na() %>% 
    summarise(
    mean_age = mean(age),
    max_age  = max(age),
    num_cases = sum(status)
    )
```

### group_by()

Group by is used to make calculations/aggregations or changes by a group.

```{r}
ebola_sl_cleaned %>% 
    drop_na() %>% 
    group_by(gender) %>% 
    summarise(
        mean_age = mean(age),
        max_age  = max(age),
        min_sample_date = min(date_of_onset),
        max_sample_date = max(date_of_onset)
    )
```

```{r}
ebola_sl_cleaned %>% 
    drop_na() %>% 
    group_by(district) %>% 
    summarise(
        mean_age = mean(age),
        min_age  = min(age),
        max_age  = max(age),
        min_sample_date = min(date_of_onset),
        max_sample_date = max(date_of_onset)
    ) %>% 
    datatable()
```

```{r}
ebola_sl_cleaned %>% 
    drop_na() %>% 
    group_by(ebola_status) %>% 
    summarise(
        mean_age = mean(age),
        min_age  = min(age),
        max_age  = max(age),
        min_sample_date = min(date_of_onset),
        max_sample_date = max(date_of_onset)
    ) %>% 
    datatable()
```

## Plotting

### Key Functions

#### ggplot()

#### aes()

#### geom\_\*()

```{r}
ggplot(ebola_sl_cleaned, 
       aes(x = status)) +
    geom_bar()
```

```{r}
ggplot(ebola_sl_cleaned, aes(x =ebola_status)) +
    geom_bar() 
```

```{r}
ggplot(ebola_sl_cleaned, aes(x =ebola_status, fill = "red")) +
    geom_bar() 
```

```{r}
ggplot(ebola_sl_cleaned, aes(x =ebola_status, fill = ebola_status,)) +
    geom_bar() 
```

### Geom_bar()

```{r}
#| echo: false

ggplot(ebola_sl, aes(x = status, 
                       group = sex, 
                       fill = sex)) +
  geom_bar(position = "dodge") +
  labs(title = "Health Status by Sex",
       x = "Health Status",
       y = "Count") +
    theme_minimal()
```

```{r}

ebola_sl_cleaned %>% 
    drop_na() %>% 
    ggplot(aes(x = ebola_status,
            group = gender,
            fill = gender)) +
  geom_bar(position = "dodge") +
  labs(title = "Health Status by Sex",
       x = "Health Status",
       y = "Count") +
    theme_minimal()
```

### Geom_histogram()

```{r}
ebola_sl_cleaned %>% 
    drop_na() %>% 
    ggplot(aes(x = age )) +
    geom_histogram(bins = 20, colour = "blue",fill = "red") +
    labs(title = "Distribution of Age",
          x ="Age",
          y = "Frequency") +
    theme_light()


```

### Geom_boxplot()

```{r}

ebola_sl_cleaned %>% 
    drop_na() %>% 
    ggplot(
       aes(x = gender, y = age, fill = sex)) +
  geom_boxplot() +
  labs(title = "Age Distribution by Sex",
       x = "Sex",
       y = "Age") +
    theme_minimal()
```

```{r}

ebola_sl_cleaned %>% 
    drop_na() %>% 
    ggplot(
       aes(x = gender, y = age, fill = gender)) +
  geom_boxplot() +
  labs(title = "Age Distribution by Sex",
       x = "Sex",
       y = "Age") +
    theme_minimal()
```

### 

#### Setting Color Uniformly

```{r}

ebola_sl_cleaned %>% 
    drop_na(sex) %>% 
    ggplot(aes(y = age, x = sex)) +
    geom_boxplot(fill = "orange")
```

#### Mapping Colors to a Variable

```{r}

ebola_sl_cleaned %>% 
    drop_na(sex) %>% 
    ggplot(aes(y = age, x = sex)) +
    geom_boxplot(aes(fill = sex))
```

#### Scaling to Customize Colors

```{r}


ebola_sl_cleaned %>% 
    drop_na() %>% 
    ggplot(aes(y = age, x = sex, fill = sex)) +
    geom_boxplot() +
    scale_fill_manual(values = c("pink", "blue")) +
    facet_wrap(~district, ncol = 2)
```

### Pie Chart

```{r}

sex_data <- ebola_sl_cleaned %>% 
    group_by(sex) %>% 
    summarise(count = n()) %>% 
    mutate(prop = count / sum(count))

sex_data
```

```{r}
sex_data %>% 
    ggplot(aes(x= "", y= prop, fill = sex)) +
    geom_bar(stat = 'identity') +
    coord_polar("y") +
    theme_void() 
```

### Scatter Plots

```{r}

ebola_sl_cleaned %>% 
    drop_na() %>% 
    filter(age >= 65) %>% 
    ggplot(aes(x = date_of_onset, 
               y = date_of_sample, 
               colour = age)) +
    geom_point(aes(size = age),
               alpha = .2) 
```

```{r}

ebola_sl_cleaned %>% 
    drop_na() %>% 
    filter(age >= 65) %>% 
    ggplot(aes(x = date_of_onset, 
               y = date_of_sample, 
               colour = age)) +
    geom_point(alpha = .2) +
    facet_wrap(~sex)
```

# Sunday Practice Session
```{r}
lbw_data <- read_excel('Low birth weight.xlsx')
lbw_data
```

```{r}
summary(lbw_data)
```

```{r}
lbw_data <- lbw_data %>% 
  rename(mother_weight = lwt,
         preterm_labor = ptl,
         hypertension = ht,
         uterine_irritability = ui,
         first_trimester_vist = ftv,
         birth_weight = bwt)

lbw_data
```

```{r}
lbw_data_new <- lbw_data %>% 
  mutate(age_group = case_when(
                age < 18 ~ "Underage",
                age >= 18 & age < 35 ~ "Young Adult",
                age >= 35 ~ "Adult",
                TRUE ~ NA_character_
                ),
         smoke_yn = if_else(smoke == 1, TRUE, FALSE )
    
  ) %>% 
  mutate(age_group = factor(age_group, 
                            levels = c("Underage",
                                       "Young Adult",
                                       "Adult")
                            )
         )

lbw_data_new
```

```{r}

lbw_data_new %>% 
  summarise(
    mean_age = mean(age),
    mean_lwt = mean(mother_weight),
    mean_bwt = mean(birth_weight),
    min_bwt = min(birth_weight), 
    max_bwt =  max(birth_weight))
```

```{r}
lbw_data_new %>%
  group_by(age_group) %>% 
  summarise(
    mean_age = mean(age) %>% round(0),
    mean_lwt = round(mean(mother_weight), 1),
    mean_bwt = mean(birth_weight),
    min_bwt = min(birth_weight), 
    max_bwt =  max(birth_weight))
```

```{r}

lbw_data_new %>%
  group_by(age_group, smoke_yn) %>% 
  summarise(
    mean_age = mean(age) %>% round(0),
    mean_lwt = round(mean(mother_weight), 1),
    mean_bwt = mean(birth_weight),
    min_bwt = min(birth_weight), 
    max_bwt =  max(birth_weight))
```

```{r}
lbw_data %>% 
  arrange(birth_weight %>% desc()) %>% 
  head(10) %>% 
  select(age, birth_weight)
```

```{r}
lbw_data_new %>% 
  ggplot(aes(x = age_group, fill = age_group)) +
    geom_bar() +
 # geom_label(aes(label = "count", y = 'count')) +
    labs(title = "Count by Age Group",
         x = "Age Group",
         y = "Frequency") +
    theme_minimal()
```

```{r}
lbw_data_new %>% 
  ggplot(aes(x = birth_weight)) +
  geom_histogram(fill = 'purple', 
                 color = "black", 
                 bins = 10) +
  theme_bw()

lbw_data_new %>% 
  ggplot(aes(x = birth_weight,)) +
  geom_histogram( fill= 'darkgreen', 
                  color = "black", 
                  bins = 20) +
  theme_classic()
```

```{r}
lbw_data_new %>% 
  ggplot(aes(x = age_group, 
             y = mother_weight, 
             fill = age_group)) +
  geom_boxplot()

```

```{r}
lbw_data_new %>% 
  ggplot(aes(x = age, y = mother_weight)) +
  geom_point(color = "green")
  
```

```{r}
lbw_data_new %>% 
  ggplot(aes(x = mother_weight, y = birth_weight)) +
  geom_point()
```

```{r}
lbw_data_new %>% 
  ggplot(aes(x = age, y = mother_weight, color = mother_weight)) +
  geom_point()
```

```{r}
lbw_data_new %>% 
  ggplot(aes(x = age, 
             y = mother_weight, 
             color = age_group)) +
  geom_point()
```